---
title: "Exploratory Data Visualization of Chocolate Sales"
author: "Cindy(Di) Chen"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    latex_engine: xelatex
---


```{r, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
library(tidyverse)
library(lubridate)
library(dplyr)
library(stringr)
library(scales)

```

### Introduction
This project analyzes a dataset of chocolate sales to identify key trends and insights about purchasing behavior. 
The dataset includes transaction dates, product categories, revenue, and geographic details. My aim is to surface 
insights around country-level performance, seasonality, and product popularity using polished and effective visualizations.

### Project Strategy: 
The goal of this analysis is to understand country-level sales performance, seasonal trends, and top products. 
Visualizations are chosen to reveal clear patterns, support decision-making for inventory planning, and ensure readability.


```{r, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
# Data Import and Cleaning

chocolate_sales <- read_csv("C:/Users/15306/Desktop/ST_537 Data Visualization/Module 8/Chocolate Sales.csv")
head(chocolate_sales)
```


```{r, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
# [Session 2: May 13, Goal: Clean date and amount columns for accurate analysis]
# Encountered issues with inconsistent date formatting; used lubridate::dmy to parse.
# Also removed "$" and "," from Amount using stringr to enable numeric conversion.

sales_clean <- chocolate_sales |>
  mutate(
    Date = dmy(Date),
    Amount = str_remove_all(Amount, "[$,]"),
    Amount = as.numeric(Amount)) |>
  drop_na() |>      
  distinct()        

```

```{r, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
summary(sales_clean)
```

## Total Amount by Country
```{r, echo=FALSE, fig.width=12, fig.height=6, out.width='100%'}
# [Session 3: May 17, Task: Refine bar chart aesthetics]
# Goal: Understand which countries contribute the most to overall revenue.
# Expectation: Anticipated USA or UK to lead in sales due to market size.
# Finding: Australia led in revenue, followed by UK, India, USA, Canada, and New Zealand.
# Action: This led to questions about regional seasonality and promotions, guiding time-series exploration.
# Visualization Notes: Added white, bold value labels inside bars and scaled Y-axis in thousands (k) for clarity.
# Used the Set2 palette for improved accessibility. Rotated x-axis labels for readability.
# Issues: The chart initially clipped bars due to plot width—fixed by increasing figure width and tweaking theme settings.
# Iterated through several layouts before finalizing this format.

ggplot(
  sales_clean |>
    group_by(Country) |>
    summarize(Amount = sum(Amount, na.rm = TRUE)) |>
    mutate(Country = fct_reorder(Country, Amount, .desc = TRUE)),
  aes(x = Country, y = Amount, fill = Country)
) +
  geom_bar(stat = "identity") +
  geom_text(
    aes(label = label_number(scale = 1e-3, suffix = "k")(Amount)),
    vjust = 1.5, color = "white", fontface = "bold", size = 3.5
  ) +
  scale_y_continuous(labels = label_number(scale = 1e-3, suffix = "k")) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Total Amount by Country", x = "Country", y = "Amount", fill = "Country") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )



```
Insight: Sales figures are highest in Australia, with the UK, India, the USA, 
Canada, and New Zealand also showing strong performance.

## Monthly Sales Trends

```{r, echo=FALSE, fig.width=12, fig.height=6, out.width='100%'}
# [Session 4: May 19, Goal: Analyze seasonality and trend patterns]
# Goal: Explore whether sales patterns fluctuate by month and tie into known holidays or promotions.
# Expectation: Expected a peak around December or Valentine’s Day.
# Finding: Highest sales occurred in January (likely post-holiday), dropped in February, rose in March, then hit a low in April.
# May and June showed a rebound, while July–August declined again.
# Action: This seasonality insight informed when to prioritize inventory of best-selling products.
# Visualization Notes: Added line and point layers for clarity, formatted x-axis by month (YYYY-MM), and angled labels to avoid overlap.
# Also add average revenue as referene line.
# Issues: Labels were initially unreadable due to long date formats — resolved with `scale_x_date` and `theme()` adjustments.


monthly_summary <- sales_clean |>
  mutate(Month = floor_date(Date, "month")) |>
  group_by(Month) |>
  summarise(Monthly_Sales = sum(Amount, na.rm = TRUE))

avg_sales <- mean(monthly_summary$Monthly_Sales)

ggplot(monthly_summary, aes(x = Month, y = Monthly_Sales)) +
  geom_line(color = "darkgreen", linewidth = 1) +
  geom_point(color = "darkgreen", size = 2) +
  geom_hline(yintercept = avg_sales, linetype = "dashed", color = "darkgrey", linewidth = 1) +
  annotate("text", x = min(monthly_summary$Month), y = avg_sales, 
           label = paste0("Avg: $", format(round(avg_sales / 1000), big.mark = ","), "K"), 
           vjust = -1, hjust = 0, color = "darkgrey", size = 3.5) +
  scale_x_date(
    date_labels = "%Y-%m",
    date_breaks = "1 month"
  ) +
  scale_y_continuous(labels = label_number(scale = 1e-3, suffix = "K")) +
  labs(
    title = "Monthly Sales Trend",
    x = "Month",
    y = "Sales ($)"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )



```

Insight: Sales peak in January, likely due to holiday demand, dip in February, 
and rise again in March. April sees the lowest sales, followed by a rebound in 
May and June before declining again in July and August

## Top 10 Selling Products

```{r, echo=FALSE, fig.width=12, fig.height=6, out.width='100%'}
# [Session 5: May 20, Goal: Identify best-selling products]
# Goal: Determine top revenue-generating products to guide inventory strategy.
# Expectation: Assumed dark chocolate variants would dominate.
# Finding: "Smooth Silky Salty" topped sales, followed by "50% Dark Bites", "White Choc", and "Peanut Butter Cubes".
# Action: These SKUs should be prioritized for restocking during sales peaks (e.g., January, May–June).
# Visualization Notes: Created a horizontal bar chart with `coord_flip()` to improve label readability.
# Issues: Initially tested different color schemes and label placements before finalizing a simple `darkorange` fill for clarity.
# Used `top_n()` and `arrange()` to isolate and order top 10 products.


top_products <- sales_clean |>
  group_by(Product) |>
  summarize(Total_Revenue = sum(Amount)) |>
  top_n(10, Total_Revenue) |>
  arrange(Total_Revenue)

ggplot(top_products, aes(x = reorder(Product, Total_Revenue), y = Total_Revenue)) +
  geom_col(fill = "darkorange") +
  geom_text(aes(label = label_number(scale = 1e-3, suffix = "K")(Total_Revenue)),
            hjust = 1.1, color = "white", fontface = "bold", size = 3.5) +
  coord_flip() +
  scale_y_continuous(labels = label_number(scale = 1e-3, suffix = "K")) +
  labs(title = "Top 10 Selling Chocolate Products",
       x = "Product", y = "Revenue") +
  theme_minimal(base_size = 12)



```
Insight: Smooth Silky Salty leads in revenue, followed by 50% Dark Bites, White 
Chocolate, Peanut Butter Cubes, Clairs, and others. These popular products should 
be prioritized for inventory planning, especially during peak sales seasons.


Session Log
Over several sessions from May 12 to May 20, I explored and cleaned the data, 
analyzed sales trends and seasonality, created and polished visualizations, and 
finalized the report ensuring reproducibility and clear narrative.
